Military mechanization as a complex adaptive system

### 2018-07-15 paper2/paper2.tex
good

\title{Military mechanization as a complex system}

@@

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is an important topic 
	in international relations scholarship and a practical concern for
	policymakers. This paper makes novel theoretical contributions by
	re-envisioning mechanization as a complex system. It also implements an
	agent-based model of military mechanization. The results of this model
	challenge the conventional wisdom that mechanization is primarily driven
	by external factors such as security environment. They imply that
	some internal factors (e.g. military culture or economic constraints)
	act as a ``brake'' on mechanization.

@@

This paper is divided into five sections. This \textbf{Introduction} gives
background on military mechanization. The \textbf{Theory} section rigorously
develops mechanization as a complex system best represented by
agent-based modeling. The \textbf{Model} section presents a hypothesis and
implements an agent-based model in Python to test that hypothesis.
\textbf{Results} are shown. Implications are explored in \textbf{Discussion}.

@@

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_j \ldots x_{P}$ covariates hypothesized to be associated with
mechanization. They conclude that, out of the
aforementioned potential causes, ``choices about
mechanization are strongly associated with a state's security environment.''
Their dataset and conclusions have guided further scholarship on military
mechanization, e.g. \citet{pilster2012democracies}, which mostly follow the same
statistical approach.

@@

Definition of complex system. 
The orthodox multivariate statistical modeling approach is insufficient because
mechanization is a complex system. Agent-based modeling (ABM) is a more 
appropriate paradigm for studying it. These three
claims---insufficiency, complexity, and ABM---are addressed in turn.

@@

\citet{sechser2010army} use observational rather than experimental
data. The assignment of treatments to subjects is not randomized, causing
\textit{selection bias}. For example, they operationalize
dyadic ``rivalry'' as whether two states recently engaged in a Militarized
Interstate Dispute (MID), and find significant association with mechanization. 
But interstate conflict is not a randomly assigned treatment. Rather, at least one state must select
itself into it \citep{fearon2002selection}. States differ in their
proclivity for conflict, based on many factors. 
Thus, ``rivalry'' is in fact
bundling together many uncontrolled covariates, casting doubt on the claimed
finding. Similar problems arise for other variables included in the model.
Natural experiments---a situation in which treatment is
known to be randomly assigned---are the typical econometric strategy to avoid selection bias. For
example, \citet{angrist2002vouchers} exploit the fact that private school
vouchers were distributed by lottery in Colombia to obtain unbiased estimates of
the effect of schooling type (private vs. public) on educational achievement.
But such situations are rare in IR: e.g., random assignment of war is an
absurd proposition. \citet{sechser2010army} lack a good control
because it probably does not exist. Reliance on selection-riddled
observational data means that econometrics are at best an incomplete
tool for studying mechanization.

\subsection{System complexity}

\subsubsection{Decomposibility}

@@

A large literature suggests that GDP per capita is associated with
regime type (e.g., democracy vs. autocracy), and causality appears to flow in
both directions. Democracy causes wealth and wealth also causes democracy \citep{helliwell1994empirical,barro1996democracy,robinson2006economic,doucouliagos2008democracy}. 
Regime type affects the probability of war: the closest thing to an empirical law in international relations is
that democracies do not fight each other
\citep{levy1988domestic}, while some kinds of autocracies are war-prone \citep{de1995war}. 
Regime type is widely argued to influence
battlefield performance in war because it affects the quality of military institutions and hence
capability of fielded forces \citep{reiter2002democracies,weeks2014dictators,talmadge2015dictator}. 
Battlefield performance affects regime type, e.g. a losing side may be subject to foreign-imposed
regime change a la 1939 France or 2003 Iraq
\citep{werner1996absolute,lo2008ensuring,downes2013forced}. \citet{sechser2010army} claim that the observed
historical probability of conflict (``rivalry'', in their terminology) affects mechanization; and claim that observed
battlefield performance (``lessons'') also affect mechanization. Mechanization surely affects
battlefield performance in war; if it did not, we would not care about it
\citep[see also][]{lyall2009rage}.
Mechanization may affect GDP per capita in that states can over-spend on defense and 
impoverish the country
\citep{cappelen1984military,ward1992sizing,abu2003government}. Deterrence theory requires that 
battlefield performance affect the probability of war in that militarily 
potent states are unlikely to be attacked. The exact truth value of these claimed relationships is
uncertain---in no small part due to the limitations of econometrics that were just addressed---but
they are generally uncontroversial statements in their respective fields.

@@

Figure 2 is the logic diagram resulting from the above. It must be emphasized that we are only scratching the 
surface of system complexity. For example, Figure 2 does not address the nature of relationships 
between covariates: linear or non-linear? subject to multi-way
interactions? necessity and sufficiency distinctions? But even as-is, Figure 2 visually conveys the essential point: \textit{the system is
not tractable to decomposition}. Ascertaining cause and effect in this system
through simple observation is impossible because of endogeneity. Moreover, controlling for
confounding variables breaks down when the controlled-for variable directly
relates to the outcome of interest \citep[``post-treatment bias'',
c.f.][]{king2010hard}. One would like to perform ``causal surgery'' \citep{pearl2009causality}, i.e. modify the value at one
of the nodes and observe the resultant system state. But, as previously covered, one generally cannot conduct
experimental trials in international relations.  A modeling paradigm that would let us do so is needed.

\subsubsection{Cognition}

There is no law of nature dictating that an increase in one state's mechanization causes an adversary to 
raise its own mechanization. It also is not the case that many individual choices aggregate into something 
akin to natural law---the logic used 
in economics to justify treating supply and demand in a physics-esque manner---because the number of
people that decide mechanization choices is small. In a personalistic
dictatorship, it might be $n=1$. Even in polities with more dispersed authority, such decisions
typically occur in the rarified and thinly populated realm of generals and cabinet ministers. 
\textit{A state's mechanization level changes because someone decides to change it}.

If one takes the previous statement to be true, then a model of military
mechanization must be at least partly a cognitive model. Some agent or set of agents perceives the outside
world; processes that information; and takes action. Orthodox studies of
mechanization incorporate cognition to some degree. For example,
\citet{sechser2010army} test for learning: whether states that fail in
counter-insurgency wars decrease their mechanization as a result of this
``anti-mechanization lesson.'' This depiction, however, fails to capture 
key phenomena.

Human beings learn not only from their own experiences, but also from observing others.
State decisionmakers are no exception. For example, the stunning success of
Coalition precision-guided munitions in the 1991 Gulf War led other states who
were not direct combatants to reform their own militaries
\citep{scobell2011chinese}. There is no guarantee that such emulation always
works out. Consider Ireland: between the 1920s and 1940s, it sought to spend a
large portion of its defense budget on building a small number of
tanks, which at that time were novel technology being implemented by the leading
European militaries.
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}. Such learning
is absent from orthodox models of military mechanization.

@@

In light of the above, agent-based modeling (ABM) is 
uniquely well-suited to study mechanization. Computer simulations provide the possibility 
to conduct controlled experiments \textit{in silico} \citep{epstein1996growing} that are not possible 
in the real world, freeing us from the tyranny of selection bias. Because the
modeler can adjust various ``dials'' and observe the outcome, ABM allows causal
inference or at least insight even in highly complex, perhaps unsolvable systems
\citep[c.f.][]{axtell2000agents}. ABM focuses on autonomous agents with
cognition, for which even relatively simple behavioral rules can generate
behavior at least as complex as mechanization
\citep{gilbert2005simulation,miller2009complex}.

@@

I hypothesize that the arguments of \citet{sechser2010army} are incomplete and
that some omitted variable or set of variables acts as a ``brake'' on 
mechanization.\footnote{By ``omitted'', I mean that Sechser and Saunders conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \citet{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes are then validated against real-world 2001 data. If
\citet{sechser2010army} are correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching my hypothesis. Proving exactly what
that missing variable or set of variables is is outside the scope of this model
and paper, though it is discussed.

@@

\subsection{Agents}

>>>>

\title{Military mechanization as a complex adaptive system}

@@

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is an important topic 
	in international relations scholarship and a practical concern for
	policymakers. This paper makes novel theoretical contributions by
	re-envisioning mechanization as a complex adaptive system. It implements an
	agent-based model of military mechanization. The results of this model
	challenge the conventional wisdom that mechanization is primarily driven
	by external factors such as security environment.

@@

This paper is divided into five sections. This \textbf{Introduction} gives
background on military mechanization. The \textbf{Theory} section rigorously
develops mechanization as a complex adaptive system best represented by
agent-based modeling. The \textbf{Model} section presents a hypothesis and
implements an agent-based model to test that hypothesis.
\textbf{Results} are shown. Implications are explored in \textbf{Discussion}.

@@

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_j \ldots x_{P}$ covariates hypothesized to be associated with
mechanization. They conclude that, out of the
potential causes mentioned above, ``choices about
mechanization are strongly associated with a state's security environment.''
States that have engaged in conflict with a highly-mechanized adversary and states
with highly mechanized neighbors tend to higher mechanization; conversely, states facing
domestic insurgency or that have lost counter-insurgency wars tend to have lower
mechanization rates. Their dataset and conclusions have guided further scholarship on military
mechanization, e.g. \citet{pilster2012democracies}, which mostly follow the same
statistical approach.

@@

The orthodox multivariate statistical modeling approach is insufficient because
mechanization is a complex adaptive system. Agent-based modeling (ABM) is a more 
appropriate paradigm for studying it. These three
claims---insufficiency, complexity, and ABM---are addressed in turn.

@@

\citet{sechser2010army} use observational rather than experimental
data. The assignment of treatments to subjects is not randomized, causing
\textit{selection bias}. For example, they operationalize
dyadic ``rivalry'' as whether two states recently engaged in a Militarized
Interstate Dispute (MID), and find significant association with mechanization. 
But interstate conflict is not a randomly assigned treatment. Rather, at least one state must select
itself into it \citep{fearon2002selection}. States differ in their
proclivity for conflict, based on many factors. 
Thus, ``rivalry'' is in fact
bundling together many uncontrolled covariates, casting doubt on the claimed
finding. Similar problems arise for other variables included in the model.
Natural experiments---a situation in which treatment is
known to be randomly assigned---are the ideal strategy to avoid selection bias. For
example, \citet{angrist2002vouchers} exploit the fact that private school
vouchers were distributed by lottery in Colombia to obtain unbiased estimates of
the effect of schooling type (private vs. public) on educational achievement.
But such situations are rare in IR: e.g., random assignment of war is an
absurd proposition. There are statistical methods for correction of selection
bias, such as Heckman correction and propensity score matching, but system
complexity prevents their application, as covered in the following sub-section.

\subsection{Complex adaptive system}

There is no single universally accepted definition of a complex adaptive system
\citep{miller2009complex}. However, two core features are
that the system is non-decomposable, and that the system features cognitive
agents. The following sections detail how mechanization meets this description.

\subsubsection{Decomposability}

@@

A large literature suggests that GDP per capita is associated with
regime type (e.g., democracy vs. autocracy), and causality appears to flow in
both directions. Democracy causes wealth and wealth also causes democracy \citep{helliwell1994empirical,barro1996democracy,robinson2006economic,doucouliagos2008democracy}. 
Regime type affects the probability of war: the closest thing to an empirical law in international relations is
that democracies do not fight each other
\citep{levy1988domestic}, while some kinds of autocracies are war-prone \citep{de1995war}. 
Regime type is widely argued to influence
battlefield performance in war because it affects the quality of military institutions and hence
capability of fielded forces \citep{reiter2002democracies,weeks2014dictators,talmadge2015dictator}. 
Battlefield performance affects regime type, e.g. a losing side may be subject to foreign-imposed
regime change a la 1939 France or 2003 Iraq
\citep{werner1996absolute,lo2008ensuring,downes2013forced}. \citet{sechser2010army} claim that the observed
historical probability of conflict (``rivalry'', in their terminology) affects mechanization; and claim that observed
battlefield performance (``lessons'') also affect mechanization. Mechanization surely affects
battlefield performance in war; if it did not, we would not care about it
\citep[see also][]{lyall2009rage}.
Mechanization may affect GDP per capita in that states can over-spend on defense and 
impoverish the country
\citep{cappelen1984military,ward1992sizing,abu2003government}. Deterrence theory requires that 
battlefield performance affect the probability of war in that militarily 
potent states are unlikely to be attacked. The exact truth value of these claimed relationships is
uncertain---in no small part due to the limitations of econometrics that were just addressed---but
they are generally mainstream statements in their respective fields.

@@

Figure 2 is the logic diagram resulting from the above. It must be emphasized that we are only scratching the 
surface of system complexity. For example, Figure 2 does not address the nature of relationships 
between covariates: linear or non-linear? subject to multi-way
interactions? necessity and sufficiency distinctions? Moreover, it only includes 
abstract variables and not individual agents (a point addressed in the following
sub-section). But even as-is, Figure 2 visually conveys the essential point: \textit{the system is
not tractable to decomposition}. We cannot isolate one component from the
greater whole.

This system complexity disallows previously-mentioned statistical techniques
for correction of selection bias. Both Heckman correction and propensity score
matching boil down to using covariates to predict (and hence control for) the chances 
of receiving treatment. For example, we might assess the effect of war on
mechanization by controlling for the factors that affect propensity for war. 
But these techniques break down if the confounding variables also directly
relate to the outcome of interest, or if the direction of causality flows both
wars. Both are true for mechanization 
(c.f. Figure 2). This ``post-treatment bias'' \citep{king2010hard} means we need
some technique other than econometrics to perform causal inference regarding
mechanization.

We would like to perform ``causal surgery'' \citep{pearl2009causality}, i.e. modify the value at one
of the nodes and observe the resultant system state. This sort of
experimentation might allow us observe the system's sensitivity (or lack thereof) to changes
in input. But, as previously covered, one generally cannot conduct
experimental trials in international relations.  A modeling paradigm that would let us do so is needed.

\subsubsection{Cognitive agents}

There is no law of nature dictating that an increase in one state's mechanization causes an adversary to 
raise its own mechanization. It also is not the case that many individual choices aggregate into something 
akin to natural law---the logic used 
in economics to justify treating supply and demand in a physics-esque manner---because the number of
people that decide mechanization choices is small. In a personalistic
dictatorship, it might be $n=1$. Even in polities with more dispersed authority, such decisions
typically occur in the rarified and thinly populated realm of generals and cabinet ministers. 
\textit{A state's mechanization level changes because someone decides to change it}. If one takes 
the previous statement to be true, then a model of military mechanization must be a model of 
cognitive agents.  Some agent or set of agents perceives the outside
world; processes that information; and takes action.

Orthodox studies of mechanization incorporate cognition to some degree. For example,
\citet{sechser2010army} test for learning: whether states that fail in
counter-insurgency wars decrease their mechanization as a result of this
``anti-mechanization lesson.'' This depiction, however, fails to capture 
key phenomena.

Human beings learn not only from their own experiences, but also from observing others.
State decisionmakers are no exception. For example, the stunning success of
Coalition precision-guided munitions in the 1991 Gulf War led other states who
were not direct combatants to reform their own militaries
\citep{scobell2011chinese}. There is no guarantee that such emulation always
works out. Consider Ireland: between the 1920s and 1940s, it sought to spend a
large portion of its defense budget on building a small number of
tanks, which at that time were novel technology being implemented by the leading
European militaries. They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}. A more
sophisticated representation of cognition that incorporates such
learning-from-others is needed.

@@

In light of the above, agent-based modeling (ABM) is 
uniquely well-suited to study mechanization. Computer simulations provide the possibility 
to conduct controlled experiments \textit{in silico} \citep{epstein1996growing} that are not possible 
in the real world, freeing us from the tyranny of selection bias. Because the
modeler can adjust various ``dials'' and observe the outcome, ABM allows causal
inference or at least insight even in highly complex, perhaps unsolvable systems
\citep[c.f.][]{axtell2000agents}. ABM focuses on autonomous agents with
cognition, for which even relatively simple behavioral rules can generate
behavior at least as complex as mechanization
\citep{gilbert2005simulation}.

@@

@@

I hypothesize that the arguments of \citet{sechser2010army} are incomplete and
that some omitted variable or set of variables acts as a ``brake'' on 
mechanization.\footnote{By ``omitted'', I mean that Sechser and Saunders conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \citet{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes are then validated against real-world 2001 data. If
\citet{sechser2010army} are correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching my hypothesis. Proving exactly what
that missing variable or set of variables is is outside the scope of this model
and paper, though it is discussed.
\subsection{Agents}
